---
title: "AVITI_07_pairwise_differential_expression_analysis"
output: html_document
date: "2025-03-18"
---

### Useful resources:
1. [Count Based RNA-seq analysis](https://kasperdanielhansen.github.io/genbioconductor/html/Count_Based_RNAseq.html)
2. [edgeR: differential analysis of sequence read count data - User’s Guide ](https://bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf)
3. [differential expression analysis of RNA-Seq experiments using the edgeR quasi-likelihood pipeline](https://doi.org/10.12688/f1000research.8987.2)

# 1. Load libraries
```{r message=FALSE}
# BiocManager::install("edgeR")
# BiocManager::install("vsn")
library(tidyverse)
library(edgeR) # v 4.2.2
```

# 2. Data importing, filtering and normalization
## 2.1 Load raw RNA-seq count data and genotype data
```{r loading data}
# load library id + genotype info
colData = read.csv(file = "path_to_file/phenodata.csv")
colData$genotype <- factor(colData$genotype,
                           levels = c("col",
                                      "xct",
                                      "prp19a",
                                      "prp19b",
                                      "prp19a_prp19b",
                                      "xct_prp19a",
                                      "xct_prp19b"))

# load raw transcript abundance data
countData=read.table("path_to_file/featurecounts_gene_no_MM.txt",
                     header=T,
                     row.names=1,
                     sep="\t")[,-1:-4]

colnames(countData)[-1] <- colData$id

# replace StringTie gene ID to AGI if gene is annotated
tmap <- read.csv("path_to_file/merged.stringtie.tmap.csv") 
tmap <- distinct(.data = tmap, qry_gene_id, .keep_all = TRUE) %>% 
  dplyr::select(-qry_id) # remove redundant StringTie qry gene ids

countData <- countData %>% 
  rownames_to_column(var = "qry_gene_id") %>% 
  left_join(tmap, by = "qry_gene_id") %>% 
  mutate(ref_gene_id = ifelse(ref_gene_id == "-", qry_gene_id, ref_gene_id)) %>%  # use StringTie id if no AGI exists (possibly a new gene) 
  mutate(ref_gene_id = ifelse(duplicated(ref_gene_id), qry_gene_id, ref_gene_id)) %>%  # use StringTie id if duplicated AGIs exist; name the second one using StringTie id instead
  filter(!duplicated(ref_gene_id)) %>%  # removed the 10 genes that still have duplicated id; checked individually that none of them are informative
  dplyr::select(ref_gene_id, everything(), -qry_gene_id) %>% 
  column_to_rownames(var = "ref_gene_id")
```

## 2.2 Filter out data with low counts
```{r expression filtering}
keep <- rowSums(countData[,-1] >= 10) >= 3
countData <- countData[keep,]
```

## 2.3 Normalization
```{r Create DGE data}
# create a numeric matrix of our count data to serve as an input to DGElist.
Data.matrix <- countData %>% 
  dplyr::select(-"Length") %>% 
  as.matrix()

DGE.data <- DGEList(counts = Data.matrix, 
                    group = colData$genotype,
                    genes = countData[, 1, drop = FALSE])

# normalize the data in our DGElist object using the Trimmed mean of M-values/TMM method
DGE.data <- calcNormFactors(DGE.data, method = "TMM")
```

# 3. Pairwise DE analysis
## 3.1 Explore data
```{r}
plotMDS(DGE.data, dim.plot = c(1,2))
# examine the samples for outliers
```

## 3.2 Build design matrix
```{r design table}
design <- model.matrix(~0+DGE.data$samples$group)
# here, the 0+ in the model formula is an instruction to include a column for each group.
colnames(design) <- levels(DGE.data$samples$group)
design
```

## 3.5 Determine differentially expressed genes
```{r DE analysis - Differential expression}
# use glmQLFit function
# compared to likelihood ratio test, Quasi-likelihood F-test provides more rigorous type I error rate control 
fit <- glmQLFit(DGE.data, design, robust = TRUE)
qlf <- glmTreat(fit, 
                contrast = c(-1, 1, 0, 0, 0, 0, 0), # adjust the contrast for the pairwise comparison (eg. xct-col)
                lfc = log2(1.2))  # fold change>1.2
summary(decideTests(qlf, adjust.method = "BH", p.value = 0.05))  # p.adjusted<0.05

plotMD(qlf) # plot log-fold change against log-counts per million, with DE genes highlighted
```

## 3.6 Save DEG list
```{r save DEG list}
result <- decideTests(qlf, adjust.method = "BH", p.value = 0.05) %>% as.data.frame()
colnames(result) <- "change"

DE_genes <- topTags(qlf, n = sum(result$change != 0))$table %>% 
  rownames_to_column(var = "AGI")
write.csv(DE_genes, "path_to_file/DEG_xct_vs_col.csv")
```

# 4. PC analysis
## 4.1 Generate log transformed TMM 
```{r TMM Data}
# log2 transformed count data for downstream processing
# do not filtered for gene significance yet
TMM_DATA <- cpm(DGE.data, log = T) %>%
  data.frame() %>%
  rownames_to_column(var = "AGI")
```

## 4.2 Calculate variance explained
```{r}
TMM_matrix <- TMM_DATA %>% 
  column_to_rownames(var="AGI") %>% 
  t()
pca <- prcomp(TMM_matrix)

library(factoextra)
fviz_eig(pca, 
         addlabels = T, 
         hjust = .5, 
         barfill = "grey",barcolor = "grey",
         geom = "bar", 
         xlab = "principal components")

PCA_data <- cbind(colData, pca$x)
PCA_data$genotype <- factor(PCA_data$genotype,
                            levels = c("col",
                                        "prp19a",
                                        "prp19b",
                                        "prp19a_prp19b",
                                        "xct",
                                        "xct_prp19a",
                                        "xct_prp19b"))
```

## 4.3 Plot data
```{r PCA plots}
mytheme <- theme_bw() +
  theme(text = element_text(size = 12)) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, face = 1)) + 
  theme(axis.text.y = element_text(face = 1)) + 
  theme(legend.text = element_text(hjust = 0)) +
  theme(axis.line = element_line(colour = "black"))

# RColorBrewer::brewer.pal(name = "Dark2", n=8)
my_color <- c("#1B9E77", "#7570B3", "#E7298A", "#D95F02", "#666666", "#E6AB02", "#A6761D")

PCA_plot1 <- PCA_data %>% 
  ggplot(mapping = aes(x=PC1, y=PC2, color=genotype)) +
  geom_point(size = 3) +
  scale_color_manual(values = my_color,  
                     name = "Genotype",
                     labels = expression("Col-0",
                                        italic("prp19a-1"),
                                        italic("prp19b-1"),
                                        italic("prp19a-1 prp19b-1"),
                                        italic("xct-2"),
                                        italic("prp19a-1 xct-2"),
                                        italic("prp19b-1 xct-2"))) +
  labs(title = "PCA for all genes", 
       x="PC1: 40.5% variance",
       y="PC2: 21.5% variance") +
  mytheme

PCA_plot1
```

# 5. GO analysis
## 5.1 Build a function to calculate fold enrichment
```{r GO analysis - setup, warning=F}
library(clusterProfiler) # version: 4.10.1
library(org.At.tair.db) # Version: 3.18.0

# this function parse character ratio to double value, such as 1/5 to 0.2
parse_ratio <- function(ratio) {
    ratio <- sub("^\\s*", "", as.character(ratio))
    ratio <- sub("\\s*$", "", ratio)
    numerator <- as.numeric(sub("/\\d+$", "", ratio))
    denominator <- as.numeric(sub("^\\d+/", "", ratio))
    return(numerator/denominator)
}
``` 

## 5.2 Run GO and build plot
```{r}
Go_genes <- read_csv("path_to_file/DEG.csv")

GO.out <- enrichGO(gene         = Go_genes$AGI,
                  universe      = rownames(countData),  # use all 20,258 expressed genes as background
                  OrgDb         = org.At.tair.db,
                  keyType       = "TAIR",
                  ont           = "BP",
                  pAdjustMethod = "fdr",
                  pvalueCutoff  = 0.05,
                  qvalueCutoff  = 0.1,
                  readable      = TRUE)

# use clusterProfiler::simplfy function to remove redundancy of enriched GO terms
GO.out_simplified <- clusterProfiler::simplify(
  x = GO.out,
  cutoff = 0.7,
  by = "pvalue",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)

GO_plot <- GO.out_simplified@result %>% 
  top_n(10, wt=-qvalue) %>%  # plot top 10 most significant GO terms
  filter(qvalue<0.05) %>%  # remove any term that has q < 0.05
  mutate(FoldEnrichment = parse_ratio(GeneRatio) / parse_ratio(BgRatio)) %>% 
  ggplot(aes(x=FoldEnrichment, 
             y=fct_reorder(Description, qvalue, .desc = T), 
             colour=qvalue, 
             size=Count)) +
  geom_point() +
  scale_color_viridis_c(guide=guide_colorbar(reverse=TRUE), direction = -1, 
                        begin = 1e-20, end = 1) +
  scale_size_continuous(range=c(2, 8)) +
  labs(title = "", 
       x="Fold Enrichment", y="", colour="q value", size="Count") +
  theme_bw() +
  theme(text = element_text(size=12))

plot(GO_plot)
```
